Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-18 paper.tex
First pass at abstract, some initial stuff on intro

\begin{abstract}
XX
\end{abstract}

\section{Introduction}
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.

The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations, increased the
detectability of recombination. By searching for genomes containing segments
from different VoC designations, multiple recombinant lineages have been
detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta~\citep{Sekizuka2022-xz}. More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of
onward transmission (more than five descendants) and monophyly ($>70\%$
bootstrap support) are eligible for a Pango lineage
designation~\citep{Rambaut2020-dw}, starting with the letter X
% FIXME: should this be a citation?
(``Pango Lineage Nomenclature: Provisional Rules for Naming Recombinant
Lineages'' 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages
\citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
candidate recombinants, RIPPLES first builds a phylogeny without assuming
recombination and then scans for long branches (i.e., carrying many mutations)
and then determines if these candidates would be better explained by
recombination. Specifically, RIPPLES exhaustively breaks each candidate
sequence into segments and attempts to find more parsimonious placements for
each segment on the phylogeny. If such placements are found, the sequence is
identified as a putative recombinant. Although it enables rapid searches for
genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
accounts for only mutations, treats recombinants \textit{post hoc}, and is an
incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
As noted by the authors, a \textit{post hoc} treatment of recombination is
possible when recombinant lineages are rare and leave few descendants. However,
this is becoming increasingly untenable with the proliferation of recombinant
lineages; for example, more than half of the sequences sampled in February 2023
are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
future evolution of SARS-CoV-2 is likely to involve multiple sequential
recombination events on top of existing recombinant lineages, creating a highly
reticulated genealogy.

%% ARGs are good for *representing* viral genealogies

It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
and affects the results of downstream analysis, e.g., inference of
selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
not account for recombination
\citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
genetic ancestries of recombining lineages can be represented by ancestral
recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
development of the ``succinct tree sequence'' data
structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
encoding of large ARGs. This representation should be well-suited to storing
large recombinant viral genealogies such as SARS-CoV-2,
which currently encompasses over 15 million sequences in the GISAID database
\citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
data structure, enables efficient manipulation of the stored genealogies and is
integrated with a suite of open-source software tools to simulate, analyse, and
infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

%% ARG inference is hard

Inferring recombinant phylogenies has been a long-standing computational challenge.
However, approaches to infer ARGs now exist that can scale to tens of
thousands of human genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
deal with levels of repeated recombination far in excess of those seen in viral
phylogenies. For example, using over 90 million variable sites
\citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
of modern and ancient human genomes, accounting for millions of recombination
events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
took advantage of a highly efficient Hidden Markov Model (HMM) implementation
of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
genomes \citep{Kelleher2019-ba}.

% What is new here?

Here, we present a preliminary version of a novel scalable methodology
(\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
and additionally takes advantage of known collection dates to infer ARGs from millions
of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
(3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
\texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
\texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
and to jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at scale.

>>>>

\begin{abstract}
Recombination is an important and ongoing feature of SARS-CoV-2 evolution,
and presents major challenges to [things important to public health].
Detection of new recombinant strains is challenging, with methods
[being a bit rubbish].
In addition, many basic approaches fundamental to the study of viral
evolution assume that recombination is negligible, in that a single
phylogenetic tree can represent the genetic ancestry of the
circulating strains. Here we present an initial version of
\texttt{sc2ts}, a method to automatically detect recombinants
in real-time and to fully integrate them into a cohesive joint
model of mutation, recombination and genetic inheritance known
as an Ancestral Recombination Graph (ARG). We infer two ARGs under
different sampling strategies, and study their properties.
One is densely sampled, and includes 1.27 million sequences
sampled up to June 30, 2021.
The second is more sparsely sampled,
consisting of 784K sequences sampled up to June 30, 2022.
We find that both ARGs are
% better words?
highly consistent with
known features of SARS-CoV-2 evolution, recovering the basic
backbone phylogeny, mutational spectra, and recapitulating
details on the majority of known recombinant lineages.
Using the well-established and fully-featured \texttt{tskit} library,
the ARGs can also be stored concisely and processed efficiently
using standard Python tools. For example, the ARG for 1.27 million
sequences---encoding the inferred reticulate ancestry,
full alignments, and extensive metadata---requires
58MB of storage,
and loads in less than a second.
%JK Not sure if we'd actually say it like this, but this is the
% basic gist I'd like to get accross[
The ability to fully integrate the effects of recombination into
downstream analyses, to quickly and automatically detect new recombinants,
and to provide an efficient and convenient platform for computation
makes \texttt{sc2ts} a promising approach, worthy of
further investment and development.
\end{abstract}

\section{Introduction}
% Recombination is an important force in SARS-CoV-2, recombinants
% have arisen and they have spread to high frequencies
Recombination via template switching is a common feature
in the evolution of coronaviruses~\citep{Graham2010-xe,De_Klerk2022-tt},
including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with adaptation to new host species and with the production of more
immune evasive variants~\citep{Graham2010-xe,De_Klerk2022-tt}.
The lack of genetic variation made recombination challenging to detect early in the SARS-CoV-2 pandemic. In late 2020, the appearance and
spread of variants of concern (VoC), designated into classes such as Alpha and
Delta which harboured multiple characteristic mutations, increased the
detectability of recombination.
More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for
recombination~\citep{Bal2022-hq}.
[FIXME say this better and add citation.]
Multiple recombinant lineages have successfully
established and spread to high frequency.

[PARA 2: recombination is hard to detect. Quick overview of state-of-the-art.
Finish with phylogeny-based method, RIPPLES.]

[PARA 3: but you can't use a phylogeny when there's recomb. Need a joint
model of mutation and recomb. ARGs, tskit, forward refs to ARG section for
details.]

[PARA 4: we present sc2ts based on these recent advances. Summarise
results, with forward refs to Results sections.]

% By searching for genomes containing segments
% from different VoC designations, multiple recombinant lineages have been
% detected, for example, between Alpha and pre-VoC \citep{Jackson2021-ik}, Alpha
% and Epsilon \citep{Wertheim2022-hj}, and Alpha and
% Delta~\citep{Sekizuka2022-xz}.

% Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
% combine specific subsets of mutations from different lineages
% \citep[e.g.][]{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}.
% This approach only identifies those inter-lineage recombination events
% resulting in the patterns investigated.
% Others have inferred recombinant
% genealogies to disentangle recombination from recurrent mutation, without
% having to pre-specify characteristic mutations for each lineage
% \citep{Ignatieva2021-rg}. More recently, \cite{Turakhia2022-it} presented a
% phylogeny-based approach (``RIPPLES'') to identify putative recombinants among
% over ten million SARS-CoV-2 genomes, without pre-specifying motifs. To find
% candidate recombinants, RIPPLES first builds a phylogeny without assuming
% recombination and then scans for long branches (i.e., carrying many mutations)
% and then determines if these candidates would be better explained by
% recombination. Specifically, RIPPLES exhaustively breaks each candidate
% sequence into segments and attempts to find more parsimonious placements for
% each segment on the phylogeny. If such placements are found, the sequence is
% identified as a putative recombinant. Although it enables rapid searches for
% genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2 phylogeny that
% accounts for only mutations, treats recombinants \textit{post hoc}, and is an
% incomplete representation of the reticulate evolutionary history of SARS-CoV-2.
% As noted by the authors, a \textit{post hoc} treatment of recombination is
% possible when recombinant lineages are rare and leave few descendants. However,
% this is becoming increasingly untenable with the proliferation of recombinant
% lineages; for example, more than half of the sequences sampled in February 2023
% are from XBB and its descendants~\citep{Chen2022-pz}. This also means that
% future evolution of SARS-CoV-2 is likely to involve multiple sequential
% recombination events on top of existing recombinant lineages, creating a highly
% reticulated genealogy.

% %% ARGs are good for *representing* viral genealogies

% It is well known that recombination distorts phylogenies \citep{Schierup2000-fg}
% and affects the results of downstream analysis, e.g., inference of
% selection~\citep{Anisimova2003-vr}. However, standard phylogenetic methods do
% not account for recombination
% \citep[e.g.][]{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}. Instead,
% genetic ancestries of recombining lineages can be represented by ancestral
% recombination graphs (ARGs)~\citep{Griffiths1981-lw,Hudson1983-properties}, which
% capture network-like phylogenetic relationships \citep{Gusfield2014-qw}. Recently,
% development of the ``succinct tree sequence'' data
% structure \citep{Kelleher2018-xc}  has enabled compact yet flexible
% encoding of large ARGs. This representation should be well-suited to storing
% large recombinant viral genealogies such as SARS-CoV-2,
% which currently encompasses over 15 million sequences in the GISAID database
% \citep{Shu2017-hp}). Moreover the \texttt{tskit} software library, which implements this
% data structure, enables efficient manipulation of the stored genealogies and is
% integrated with a suite of open-source software tools to simulate, analyse, and
% infer them \citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

% %% ARG inference is hard

% Inferring recombinant phylogenies has been a long-standing computational challenge.
% However, approaches to infer ARGs now exist that can scale to tens of
% thousands of human genomes and beyond
% \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
% deal with levels of repeated recombination far in excess of those seen in viral
% phylogenies. For example, using over 90 million variable sites
% \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
% of modern and ancient human genomes, accounting for millions of recombination
% events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
% took advantage of a highly efficient Hidden Markov Model (HMM) implementation
% of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
% genomes \citep{Kelleher2019-ba}.

% % What is new here?

% Here, we present a preliminary version of a novel scalable methodology
% (\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
% and additionally takes advantage of known collection dates to infer ARGs from millions
% of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
% we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
% recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
% (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
% of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
% \texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
% \texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
% and to jointly model recombinational and mutational processes shaping the genomic
% diversity of SARS-CoV-2 at scale.
